---
title: '95868 Homework 1'
author: "Lara Haase (Lhaase)"
output: 
  html_document:
    fig_width: 8
    fig_height: 10
---

#### Instructions 

Submit this Rmd file on blackboard. Don't submit additional files. Don't forget to put your name above!!

Fill in the missing code chunks to answer the following questions. You should use a reasonable coding style and comment your code so that it is clear what you are doing. As a rough rule of thumb, expect to add multiple comments to each missing code chunk.

If the question asks for a short sentence description, you can type this into the file directly.

Here are two recommended ways that you can report your answer:

**Example Question 1:** In the `cars` data set, what is the average value of the `speed` column?

```{r}
# take average of mpg$speed, save as variable
avg.speed = mean(cars$speed)
```

**Answer:** The average value was `r avg.speed`

**Example Question 1:** In the `ChickWeight` dataset, what was the average weight of the chickens grouped by each diet type?

```{r}
library(plyr)
library(knitr) # for the kable() function, which prints data frames as tables:

# group the chickens by diet and take average weight
avg.weights = ddply(ChickWeight, "Diet", summarize, avg.weight = mean(weight))
# display the data frame as an easy-to-read table, with 3 significant digits
kable(avg.weights, digits = 3, col.names = c('Diet', 'Average Weight'))

# or you can display more primitively if the data frame is small
avg.weights
```


#### Questions 

**Note:** We will assume that the data file `hw1_data.csv` is in your working directory.

**Problem 1:** 

Load the CSV file `hw1_data.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable called `expenditure.small`. 

Hint: You may want to look at the CSV file using another program such as excel to see if the first row is a header. 

```{r}
# Read data into environment and save as variable "expenditure.small"
# default settings read the first line as a header
expenditure.small<- read.csv('hw1_data.csv')


```

**Problem 2:**

a) How many records are in the data set? 


b) The column `home.type` describes the type of home belonging to each respondent. What are the different levels that this column is allowed to take?
```{r}
# We don't need this code section -- the answers are all inline (see below):
```

**ANSWER:** 

(a) `r nrow(expenditure.small)`
(b) `r levels(expenditure.small$home.type)`

**Problem 3:**

What fraction of records in the data set belong to each `home.type`? In your answer, order the home types by how many records they have (largest first).

```{r}
# Fill in, with comments added:

#group with ddply function, by "home.type" column 
#fraction_of_total = find length(# of records) of each level in home.type then divide by 4000 (the total number of households)
#save in the variable "type.percent"
type.percent <- ddply(expenditure.small, c("home.type"), summarize, 
               fraction_of_total = length(home.type)/4000, .drop = F) 

# Sort type.percent in descending order by fraction_of_total (of home.type)
type.percent<- type.percent[order(type.percent$fraction_of_total, decreasing = TRUE),]

print(type.percent) # print output
```


**Problem 4:**

Change the `expenditure` data frame so that `home.type` column has only 3 coded types: 

1. single family detached homes
2. apartments or flats
3. other (everything else)

Print the number of records in each `home.type` as a table. 

```{r}

# use mapvalues to relabel all levels as "other" (except "single family detached" and "apartment or flat")  
expenditure.small$home.type = mapvalues(expenditure.small$home.type, 
                                from=c("3-plex or 4-plex", "dormitory", "duplex", "end row or end townhouse", "garden", "high-rise", "mobile home or trailer", "row or townhouse inner unit"), to= c("other", "other", "other", "other", "other", "other", "other", "other"))

# Print number of records in each home.type as a table
table(expenditure.small$home.type)
```

**Problem 5a:**

Create a subset of the `expenditure` data set consisting of just those households whose income rank is above 95%. We will call these the "rich" households. 

```{r}
# subset the data.frame expenditure.small with all rows "which" have an income.rank greater than 0.95
# save this subset as varibale "rich"
rich<- expenditure.small[which(expenditure.small$income.rank>0.95),]

```

(you don't have to print the result, we'll use it in the next part)

**Problem 5b:**

What fraction of the households in the `expenditure` data set live in the Midwest, Northeast, South, West, and NA regions of the US? In comparison, how does your answer change if you look only at the subset of the data consisting of those households whose income rank is above 95% (from problem 5a)? Give both results, and then describe in words how they are different. 

```{r}
#group data with ddply function, by "region" column 
#fraction_of_total = find length(# of records) of each level in region then divide by 4000 (the total number of households)
#save in the variable "region.percent"
region.percent <- ddply(expenditure.small, c("region"), summarize, 
               fraction_of_total = length(region)/4000, .drop = F)

print(region.percent) #print output table
```

```{r}

#group "data"rich" subset with ddply function, by "region" column 
#fraction_of_total = find length(# of records) of each level in region and divide by the length of the whole subset (# of records in "rich")
#save in the variable "rich.region"
rich.region <- ddply(rich, c("region"), summarize, 
               fraction_of_total = length(region)/length(rich[,"region"]), .drop = F)

print(rich.region) #print output

```

How are the distributions different?

**While there is a non-uniform distribution of households across the different regions of the US, ranging from 18% (in the Northeast) to 36% (in the South), the distribution of "rich" households seems to be more even, only ranging from 21% (in the Midwest) to 28% (in the South)**




Discussion: In this question, we are comparing the regional distribution of the entire dataset, vs the regional distribution of just the people who are in the top 5% of income. Sometimes students are confused about how to use percentages to compare two different groups When using percentages to compare groups, the percentages should sum to 100% within each group.

Discussion: Often times you may feel that the instructions are ambiguous and a plot or table can be constructed multiple ways. In these cases, **make the plot each way
before asking me for clarification. Then try to figure out on your own which way makes more sense for understanding the data.** After this, if you are still unclear, let me know. There will **always** be ambiguity in going from english instructions to code -- in fact, in the workplace your client might not exactly know which method will be best for their data (after all, they are asking you to analyze it), and it will be up to you to figure this out on your own.